A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-10-09 paper.tex
more comments

% error monad
\newcommand{\just}[1]{\ensuremath{#1}}
\newcommand{\error}[1]{\ensuremath{\Uparrow#1}}

@@

The syntax \error{l} occurs several times in the syntax definition. We shall 
view those nonterminals that have an \error{l} case as error monads.
That is, they raise an error with the error message $l$.
The monadic \texttt{let} is defined as follows:
\[
\begin{array}{rcll}
    (\mlet{x}{(\error{l})}{e}) & = & \error{l} \\
    (\mlet{x}{a}{e}) & = & [a/x]e & \text{otherwise} \\ 
\end{array}
\]
In Agda code we use the monad bind operator instead.
% We use 
% the symbol \mbind{} as the monad bind operator, which is defined as follows:
% \[
% \begin{array}{rcll}
% (\error{l}) \mbind f & = & \error{l} \\
%           a \mbind f & = & f(a)      & \text{otherwise}
% \end{array}
% \]
\todo[inline]{JS: why not use monadic let instead? That avoids
  the need to also use meta-lambdas, which you would need to
  explain, e.g., in the transition rule that uses applyCast.}
\todo[inline]{KC: 1. I will try to switch to monadic let. As you pointed out
when we wrote the WGT paper, let is more friendly to readers who are not familiar
with monads.
2. Do you imply that monadic let doesn't need to be explained?
}
\todo[inline]{JS: No, you should still define monadic let.}
\todo[inline]{KC: KC, you need to replace all uses of monad bind to let.}

@@

The transition relation \reduceto{s}{s} invokes the cast application
function, noted as $\applyCast{v}{\ccast{S}{l}{T}}$, which is also defined
by specify blame strategies (Fig.~\ref{fig:blame-strategy-LazyD} and
Fig.~\ref{fig:blame-strategy-LazyUD}).
\todo[inline]{JS: be consistent with the font used for applyCast. Sometimes
  you use roman and sometimes you used italic. Also, be consistent
  for all the metafunctions (eval, observe, etc.).}
\todo[inline]{KC: Fixed}
%
When evaluating a cast expression, the machine moves the cast to the
continuation and evaluates the inner expression.
%
To cast a value, the machine invokes applyCast on the value. The cast
result is either a value or an error (\error{l}). If the cast
succeeds, the machine returns the value to the next continuation.  If
the cast fails, the machine halts with the blame label.
%
To apply a casted function, the machine first casts the argument $v$, then
applies the inner function $u$ to the casted argument, and finally
casts the return value.
%
To project the first element of a casted pair, the machine projects
the first element of the inner pair $u$ and cast it.
%
Projecting the second element of a casted pair is similar.

@@

\todo[inline]{JS: what does ``them'' refer to in the previous sentence?
  Please replace with a specific noun. Also, why should the reader do
  the comparison? What are they suppose to notice?}
\todo[inline]{JS: say where the following equations come from.
  They are the equations you needed for the proof of equivalence,
  right?}
\todo[inline]{KC: [Fixed.}

\begin{definition}[Lazy D Cast ADT]
	\label{def:LazyD-CastADT}
	A Cast ADT is a Lazy D Cast ADT if:
	\begin{enumerate}
		\item $\denote{\translate{\ccast{\Tdyn}{l}{\Tdyn}}}(v) = v$
		\item $\denote{\translate{\ccast{P}{l}{\Tdyn}}}(v) = \vdyn{P}{v}$
		\item $\denote{\translate{\ccast{\Tdyn}{l}{Q}}}(\vdyn{P}{u}) = 
		\denote{\translate{\ccast{P}{l}{Q}}}(u)$
		\item $\denote{\translate{\ccast{\Pbool}{l}{\Pbool}}}(v) = v$
		\item 
		$\denote{\translate{\ccast{\Pfunc{S_1}{T_1}}{l}{\Pfunc{S_2}{T_2}}}}(
			\ecast{\vfunc{x}{S}{e}{E}}{\Pfunc{c}{d}})$\\
		$=
		\ecast{\vfunc{x}{S}{e}{E}}{
			\Pfunc{
				(\compose{\translate{\ccast{S_2}{\bar{l}}{S_1}}}{c})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item 
		$\denote{\translate{\ccast{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}(
		\ecast{\vcons{v_1}{v_2}}{\Pprod{c}{d}})$\\
		$=
		\ecast{\vcons{v_1}{v_2}}{
			\Pprod{
				(\compose{c}{\translate{\ccast{S_1}{l}{S_2}}})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item $\denote{\translate{\ccast{P}{l}{Q}}}(v) = \error{l}$\\
			if $\notshallowlyconsistent{P}{Q}$
	\end{enumerate}
\end{definition}

@@

\begin{definition}[Lazy UD Cast ADT]
	\label{def:LazyUD-CastADT}
	A Cast is a Lazy UD Cast if:
	\begin{enumerate}
	\item $\denote{\translate{\ccast{\Tdyn}{l}{\Tdyn}}}(v) = v$
	\item $\denote{\translate{\ccast{G}{l}{\Tdyn}}}(v) = \vdyn{G}{v}$
	\item $\denote{\translate{\ccast{P}{l}{\Tdyn}}}(v) = 
		\denote{\translate{\ccast{P}{l}{G}}}(v) \mbind 
		\denote{\ccast{G}{l}{\Tdyn}}$\\
		if $G \sim P$ and $G \neq P$	
	\item $\denote{\translate{\ccast{\Tdyn}{l}{G}}}(\vdyn{G}{u}) = u$
	\item $\denote{\translate{\ccast{\Tdyn}{l}{H}}}(\vdyn{G}{u}) = \error{l}$\\
		if $G \neq H$
	\item $\denote{\translate{\ccast{\Tdyn}{l}{Q}}}(v) = 
		\denote{\translate{\ccast{\Tdyn}{l}{H}}}(v) \mbind
		\denote{\translate{\ccast{H}{l}{Q}}}$\\
		if $H \sim Q$ and $H \neq Q$
	\item $\denote{\translate{\ccast{\Pbool}{l}{\Pbool}}}(v) = v$
		\item 
		$\denote{\translate{\ccast{\Pfunc{S_1}{T_1}}{l}{\Pfunc{S_2}{T_2}}}}(
		\ecast{\vfunc{x}{S}{e}{E}}{\Pfunc{c}{d}})$\\
		$=
		\ecast{\vfunc{x}{S}{e}{E}}{
			\Pfunc{
				(\compose{\translate{\ccast{S_2}{\bar{l}}{S_1}}}{c})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item 
		$\denote{\translate{\ccast{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}(
		\ecast{\vcons{v_1}{v_2}}{\Pprod{c}{d}})$\\
		$=
		\ecast{\vcons{v_1}{v_2}}{
			\Pprod{
				(\compose{c}{\translate{\ccast{S_1}{l}{S_2}}})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item $\denote{\translate{\ccast{P}{l}{Q}}}(v) = \error{l}$\\
			if $\notshallowlyconsistent{P}{Q}$
	\end{enumerate}
\end{definition}

@@

\todo[inline]{KC: (About the following paragraph) Should we mention KC and Deyaa as citation?
  (something like in private communication with KC and Deyaa)}
\todo[inline]{JS: Yes}
\todo[inline]{KC: Could you show me how to do this? I don't know exactly
how to write this kind of citations.}

The \compose{c}{c} operator is the key to compressing coercions and
maintaining space-efficiency. The composition operator for Coercions
in Normal Form requires the target type of the first coercion to match
the source type of the second. This requirement, however, imposes
considerable difficulty in the termination proof of the composition
for \LD\ Coercions in Normal Form, reported by Kuang-Chen Lu and
Deyaaeldeen Almahallawi. So it is useful to relax this restriction for
the \LD\ tracking strategy.
%
We shall need an optional source cast, written $g$, that ranges over
$\gnone$ or $\gsome{S}{l}{T}$, to maintain the gap between the two
hypercoercions to be composed. The source cast is mandatory when the
target type of the first hypercoercion does not match the source type
of the second.  Then we write the composition of hypercoercions as
\composegap{c_1}{g}{c_2}.
%  
When both $c_1$ and $c_2$ are \hcci, their composition is also \hcci. 
%
When the first hypercoercion is the \hcci, the output must start with
a projection, in which case we invoke the helper function
\textit{mkProj}. In \textit{mkProj}, if the head of $c_2$ is a
projection, the result is $c_2$.  Otherwise, we need a label to build
the projection. Since the head of $c_2$ is a no-op, its source type
must be a pre-type. Thus we know $g$ must be a source cast and we use
it to construct the projection.
%
Going back to $\composegap{c}{g}{c}$, when the second hypercoercion is the 
\hcci, the output must end with an injection, in which case we invoke the 
helper function \textit{mkInj}.
%
Finally, when both inputs are of the form \hccc{h}{m}{t}, we compose the middle 
with an updated gap.
%
The second and third cases of \mkgap{t}{g}{h} is interesting, where we choose 
the label in projection to be responsible for potential inconsistency.

@@

\todo[inline]{KC: I need a better name for \compatibletailhead{t}{h},
any suggestion from Jeremy?}

>>>>

% error monad
\newcommand{\just}[1]{\ensuremath{#1}}
\newcommand{\error}[1]{\ensuremath{\mathop{\Uparrow}#1}}

@@

The syntax \error{l} occurs several times in the syntax definition. We shall 
view those nonterminals that have an \error{l} case as error monads.
That is, they raise an error with the error message $l$.
The monadic \texttt{let} is defined as follows:
\[
\begin{array}{rcll}
  (\mlet{x}{r}{e}) & = &
  \begin{cases}
    \error{l} & \text{if } r = \error{l} \\
    [r/x]e & \text{otherwise}
  \end{cases}
    %% (\mlet{x}{(\error{l})}{e}) & = & \error{l} \\
    %% (\mlet{x}{a}{e}) & = & [a/x]e & \text{otherwise} \\ 
\end{array}
\]
In the Agda code we use the monad bind operator instead.
% We use 
% the symbol \mbind{} as the monad bind operator, which is defined as follows:
% \[
% \begin{array}{rcll}
% (\error{l}) \mbind f & = & \error{l} \\
%           a \mbind f & = & f(a)      & \text{otherwise}
% \end{array}
% \]
\todo[inline]{KC: KC, you need to replace all uses of monad bind to let.}

@@

The transition relation \reduceto{s}{s} invokes the cast application
function, noted as $\applyCast{v}{\ccast{S}{l}{T}}$, which is also defined
by specify blame strategies (Fig.~\ref{fig:blame-strategy-LazyD} and
Fig.~\ref{fig:blame-strategy-LazyUD}).
%
When evaluating a cast expression, the machine moves the cast to the
continuation and evaluates the inner expression.
%
To cast a value, the machine invokes applyCast on the value. The cast
result is either a value or an error (\error{l}). If the cast
succeeds, the machine returns the value to the next continuation.  If
the cast fails, the machine halts with the blame label.
%
To apply a casted function, the machine first casts the argument $v$, then
applies the inner function $u$ to the casted argument, and finally
casts the return value.
%
To project the first element of a casted pair, the machine projects
the first element of the inner pair $u$ and cast it.
%
Projecting the second element of a casted pair is similar.

@@

\begin{definition}[Lazy D Cast ADT]
	\label{def:LazyD-CastADT}
	A Cast ADT is a Lazy D Cast ADT if:
	\begin{enumerate}
		\item $\denote{\translate{\ccast{\Tdyn}{l}{\Tdyn}}}(v) = v$
		\item $\denote{\translate{\ccast{P}{l}{\Tdyn}}}(v) = \vdyn{P}{v}$
		\item $\denote{\translate{\ccast{\Tdyn}{l}{Q}}}(\vdyn{P}{u}) = 
		\denote{\translate{\ccast{P}{l}{Q}}}(u)$
		\item $\denote{\translate{\ccast{\Pbool}{l}{\Pbool}}}(v) = v$
		\item 
		$\denote{\translate{\ccast{\Pfunc{S_1}{T_1}}{l}{\Pfunc{S_2}{T_2}}}}(
			\ecast{\vfunc{x}{S}{e}{E}}{\Pfunc{c}{d}})$\\
		$=
		\ecast{\vfunc{x}{S}{e}{E}}{
			\Pfunc{
				(\compose{\translate{\ccast{S_2}{\bar{l}}{S_1}}}{c})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item 
		$\denote{\translate{\ccast{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}(
		\ecast{\vcons{v_1}{v_2}}{\Pprod{c}{d}})$\\
		$=
		\ecast{\vcons{v_1}{v_2}}{
			\Pprod{
				(\compose{c}{\translate{\ccast{S_1}{l}{S_2}}})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item $\denote{\translate{\ccast{P}{l}{Q}}}(v) = \error{l}$ \qquad
			if $\notshallowlyconsistent{P}{Q}$
	\end{enumerate}
\end{definition}

@@

\begin{definition}[Lazy UD Cast ADT]
	\label{def:LazyUD-CastADT}
	A Cast is a Lazy UD Cast if:
	\begin{enumerate}
	\item $\denote{\translate{\ccast{\Tdyn}{l}{\Tdyn}}}(v) = v$
	\item $\denote{\translate{\ccast{G}{l}{\Tdyn}}}(v) = \vdyn{G}{v}$
	\item $\denote{\translate{\ccast{P}{l}{\Tdyn}}}(v) = 
		\denote{\translate{\ccast{P}{l}{G}}}(v) \mbind 
		\denote{\ccast{G}{l}{\Tdyn}}$\\
		if $G \sim P$ and $G \neq P$	
	\item $\denote{\translate{\ccast{\Tdyn}{l}{G}}}(\vdyn{G}{u}) = u$
	\item $\denote{\translate{\ccast{\Tdyn}{l}{H}}}(\vdyn{G}{u}) = \error{l}$ \qquad
		if $G \neq H$
	\item $\denote{\translate{\ccast{\Tdyn}{l}{Q}}}(v) = 
		\denote{\translate{\ccast{\Tdyn}{l}{H}}}(v) \mbind
		\denote{\translate{\ccast{H}{l}{Q}}}$\\
		if $H \sim Q$ and $H \neq Q$
	\item $\denote{\translate{\ccast{\Pbool}{l}{\Pbool}}}(v) = v$
		\item 
		$\denote{\translate{\ccast{\Pfunc{S_1}{T_1}}{l}{\Pfunc{S_2}{T_2}}}}(
		\ecast{\vfunc{x}{S}{e}{E}}{\Pfunc{c}{d}})$\\
		$=
		\ecast{\vfunc{x}{S}{e}{E}}{
			\Pfunc{
				(\compose{\translate{\ccast{S_2}{\bar{l}}{S_1}}}{c})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item 
		$\denote{\translate{\ccast{\Pprod{S_1}{T_1}}{l}{\Pprod{S_2}{T_2}}}}(
		\ecast{\vcons{v_1}{v_2}}{\Pprod{c}{d}})$\\
		$=
		\ecast{\vcons{v_1}{v_2}}{
			\Pprod{
				(\compose{c}{\translate{\ccast{S_1}{l}{S_2}}})}{
				(\compose{d}{\translate{\ccast{T_1}{l}{T_2}}})}}$
		\item $\denote{\translate{\ccast{P}{l}{Q}}}(v) = \error{l}$\qquad
			if $\notshallowlyconsistent{P}{Q}$
	\end{enumerate}
\end{definition}

@@

The \compose{c}{c} operator is the key to compressing coercions and
maintaining space-efficiency. The composition operator for Coercions
in Normal Form requires the target type of the first coercion to match
the source type of the second. This requirement, however, imposes
considerable difficulty in the termination proof of the composition
for \LD\ Coercions in Normal Form~\citep{almahallawi19}. So it is
useful to relax this restriction for the \LD\ tracking strategy.
%
We shall need an optional source cast, written $g$, that ranges over
$\gnone$ or $\gsome{S}{l}{T}$, to maintain the gap between the two
hypercoercions to be composed. The source cast is mandatory when the
target type of the first hypercoercion does not match the source type
of the second.  Then we write the composition of hypercoercions as
\composegap{c_1}{g}{c_2}.
%  
When both $c_1$ and $c_2$ are \hcci, their composition is also \hcci. 
%
When the first hypercoercion is the \hcci, the output must start with
a projection, in which case we invoke the helper function
\textit{mkProj}. In \textit{mkProj}, if the head of $c_2$ is a
projection, the result is $c_2$.  Otherwise, we need a label to build
the projection. Since the head of $c_2$ is a no-op, its source type
must be a pre-type. Thus we know $g$ must be a source cast and we use
it to construct the projection.
%
Going back to $\composegap{c}{g}{c}$, when the second hypercoercion is the 
\hcci, the output must end with an injection, in which case we invoke the 
helper function \textit{mkInj}.
%
Finally, when both inputs are of the form \hccc{h}{m}{t}, we compose the middle 
with an updated gap.
%
The second and third cases of \mkgap{t}{g}{h} is interesting, where we choose 
the label in projection to be responsible for potential inconsistency.

@@

\todo[inline]{KC: I need a better name for \compatibletailhead{t}{h},
  any suggestion from Jeremy?}
\todo[inline]{JS: how about $\sim$}
